Decompositions of congruence subgroups \\ of Chevalley groups

### 2018-02-16 paper.tex
More fixes

\section{Introduction}\label{sec:intro}
This paper studies decompositions of Chevalley groups over commutative rings (see e.\,g.~\cite{VP} and references therein for the introduction to the theory of Chevalley groups over rings).
Recall that classical decompositions of Chevalley groups over fields such as e.\,g. Bruhat decomposition do not generalize to groups over more general classes of rings.
Nevertheless, Chevalley groups over rings of {\it finite stable rank} do admit several remarkable ``parabolic decompositions'', i.\,e. decompositions formulated in terms of parabolic and unipotent subgroups.
Main interest in these decompositions comes mainly from the study of stability problems for $\K_1$-function modeled on Chevalley groups (see e.\,g.~\cite{St78, Si13}).

The first goal of this paper is to prove analogues of these decompositions for congruence subgroups of Chevalley groups of level $I \trianglelefteq R$
 while making ring-theoretic assumptions only about the ideal $I$ and not the ring $R$ itself.
One of our main results is the following theorem, which is, essentially, a version of Dennis--Vaserstein decomposition for the relative elementary group $\E(\Phi, R, I)$
 (cf.~\cite[Theorems~2.5 and~4.1]{St78},~\cite[Theorem~1.2]{Si13}).
\begin{thm}\label{thm:DennisVaserstein}
Let $\Phi \neq \rE_8$ be an irreducible root system of rank $\ell\geqslant 2$ and $\{ r, s \}$ be a pair of distinct endnodes of the Dynkin diagram of $\Phi$.
Denote by $d$ the distance between $r$ and $s$ on the Dynkin diagram.
Assume that $\Phi$, $I$ and $\{r, s\}$ satisfy either of the following assumptions:
\begin{thmlist}
 \item $\sr(I) \leqslant d$ for $\Phi$ classical or $\Phi=\rG_2, \rF_4$;
 \item $\sr(I) \leqslant d$ for $\Phi=\rE_6,\rE_7$ with $\{r, s\} = \{2, \ell\}$;
 \item $\asr(I) \leqslant d$ for $\Phi=\rE_6,\rE_7$ with $\{r, s\} = \{1, \ell \}$, 
\end{thmlist}
Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$
\end{thm}
The notation for the groups and ring-theoretic invariants appearing in the above statement is introduced in Sections~\ref{sec:preliminaries} and~\ref{sec:stability-conditions}.

Our further results in this direction are relative analogues of Gauss and Bass--Kolster decompositions (see~\cref{thm:srRI1,thm:BassKolster}).
%TODO: For what?
Another goal of this paper is to apply parabolic decompositions for obtaining the estimates of the width of relative Chevalley groups in terms of certain generating sets.
For example, in~\cref{thm:width} we obtain upper estimates on the width of $\E(\Phi, R, I)$ with respect to Tits--Vaserstein generators for several small-dimensional rings. 
Also, as a corollary of Bass--Kolster decomposition we obtain upper estimates of the number of factors in the presentation of a classical Chevalley group
 as a product of its subsystem subgroups $\SL(2, R, I)$, see~\cref{thm:SL2width}.
Moreover, as a corollary of~\cref{thm:DennisVaserstein} we obtain the following theorem, which is inspired by the main result of~\cite{Nik07}.
\begin{thm}\label{thm:spin-sln-prod}
Assume that $\sr(I) \leqslant 2$. Then the group $\Epin(2\ell, R, I)=\E(\rD_\ell, R, I)$ is a product of at most $9$ regularly embedded subgroups of type $\rA_{\ell-1}$.
\end{thm}

@@

Let $\G(\Phi, R)$ be the simply-connected Chevalley group of type $\Phi$ over an arbitrary commutative ring $R$ and let $\E(\Phi, R)$ be its elementary subgroup,
i.\,e. the subgroup generated by the elementary root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in R$, see~\cite{VP, St78, S}.
For an ideal $I \trianglelefteq R$ we denote by $\G(\Phi, R, I)$ the \emph{principal congruence subgroup} of level $I$.

The \emph{relative elementary Chevalley subgroup} $\E(\Phi, R, I) \leqslant \G(\Phi, R, I)$ is defined as the normal closure in $\E(\Phi, R)$ of the subgroup $\E(\Phi, I)$ generated by the set $\mathcal{X} = \{x_\alpha(\xi) \mid \alpha\in \Phi, \ \xi \in I \}$.

Set $z_\alpha(s, \xi) = x_{\alpha}(s)^{x_{-\alpha}(\xi)}$, $s\in I$, $\xi\in R$.
It is a classical result of J.~Tits and L.~Vaserstein (see~\cite{Tits76}, \cite[Theorem~2]{Va86}) that the relative elementary subgroup $\E(\Phi, R, I)$ is generated by the set of elements $\mathcal{Z}(\Phi)$.
Here for a subset of roots $S\subseteq\Phi$ we denote by $\mathcal{Z}(S)$ the set
\[ \mathcal{Z}(S) =  \mathcal{X} \cup \{ z_\alpha(s, \xi) \mid s\in I, \ \xi \in R, \ \alpha \in S\}.\]
\begin{lemma}[{\cite[Corollary~3.3]{S}}]\label{lemma:Stepanov-ideal}
Let $\Phi$ be a root system of rank $\geqslant2$, let $R$ be a commutative ring and $I\trianglelefteq R$ be its ideal.
If $\Phi\neq\rC_\ell$ then $\E\left(\Phi, R, I^2\right)\leqslant\E(\Phi, I)$, otherwise $\E\left(\Phi, R, II^{\indexbox{2}}\right)\leqslant\E(\Phi, I)$.
\end{lemma}
Here $I^{\indexbox{2}}$ denotes the ideal generated by the squares $a^2$, where $a\in I$.

For $\varepsilon\in R^*$ set $w_\alpha(\varepsilon) = x_\alpha(\varepsilon) x_{-\alpha}(-\varepsilon^{-1}) x_{\alpha}(\varepsilon).$
If $\rk(\Phi)\geqslant 2$ the following relation holds:
\begin{equation}\label{rel:R3}
w_\alpha(\varepsilon) x_{\beta}(\xi) w_\alpha(\varepsilon)^{-1} =
x_{\sigma_{\alpha}\beta} \left(\eta_{\alpha, \beta}\cdot\varepsilon^{-\langle\beta, \alpha \rangle}\xi\right), \quad \varepsilon\in R^*, \ \xi\in R.
\end{equation}
where $\eta_{\alpha, \beta}=\pm 1$. The coefficients $\eta_{\alpha, \beta}$ can be expressed in terms of the structure constants of the corresponding Lie algebra (see~\cite[\S13]{VP}).
For a reductive subset $\Delta \subseteq \Phi$ denote by $\widetilde{W}(\Delta)$ the \emph{extended Weyl group}, i.\,e. the subgroup of $\E(\Phi, R)$ generated by all $w_{\alpha}(1)$, $\alpha \in \Delta$.
\begin{lemma} \label{lemma:weylfacts} Let $\Phi$ be an irreducible root system and let $I$ be an ideal of $R$. 
\begin{lemlist}
\item \label{item-trans1} For every two roots $\alpha, \beta \in \Phi$ of the same length there exists $w \in \widetilde{W}(\Phi)$ such that $X_{\alpha}(I)^w = X_\beta(I)$.
\item \label{item-trans2} Let $\alpha_s\in \Pi$ be a fundamental root and $\alpha, \beta \in \Sigma^\pm_s$ be two roots having the same length and $s$-shape.
 Then there exists $w\in \widetilde{W}(\Delta_s)$ such that $X_\alpha(I)^w = X_\beta(I)$. \end{lemlist}
\end{lemma}
\begin{proof}
The first assertion follows from \eqref{rel:R3} and the well-known fact that $W(\Phi)$ acts transitively on the set of roots of the same length.
The second assertion follows from \eqref{rel:R3} and \cref{lemma:abs}.
\end{proof}

@@

From Levi decomposition~\eqref{rel:Levi-decomp} we obtain the following equality of subsets of $\E(\Phi, R, I)$:
\begin{multline*}
\U(\Phi^+, I)\cdot \U(\Phi^-, I) \cdot \E(\Delta_r, R, I) \cdot \EP_s(R, I) = \\
= \U(\Sigma_r^+, I)\cdot \U(\Sigma^-_r, I) \cdot \E(\Delta_r, R, I) \cdot \EP_s(R, I) = \hspace{5em} \\
\hspace{5em} = \EP_r(R, I) \cdot \E(\Delta_s, R, I) \cdot \U(\Sigma_s^-, I)\cdot \U(\Sigma_s^+, I) = \\
= \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).
\end{multline*}
Denote the above subset by $A_{rs}$. 
Notice that if the equality $A_{rs} = \E(\Phi, R, I)$ implies the ``symmetric'' equality $A_{sr} = \E(\Phi, R, I)$, 
 therefore it suffices to consider only the possibilities for $\Phi$, $s$, $r$ listed in~\cref{table:dv-reps} below.

\begin{table}[htb]
\[\begin{array}{l @{\qquad} l @{\qquad} c @{\quad} c @{\quad} c @{\qquad} c @{\qquad} c @{\qquad} c}
\Phi                                  & s    &r      & |\Lambda(\pi)| & \text{type of $\pi$} & \text{type of $\Delta_r$} & |\Lambda(\pi')|& |\Gamma|  \\ \hline\vphantom{\Bigl(}
\rA_\ell, \ \ell\geqslant 2           & 1    &\ell   & \ell+1         & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rB_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell+1        & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rC_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\
\rD_\ell, \ \ell\geqslant 4           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 2  \\ 
\rD_\ell, \ \ell\geqslant 4           & \ell &\ell-1 & 2^{\ell-1}     & \text{half-spinor}   & \rA_{\ell-1}              & \ell           & \ell-2  \\
\rE_\ell, \ \ell=6, 7                 & \ell &2      & 27, 56         & \text{minimal}       & \rA_{\ell-1}              & \ell           & 3       \\ 
\rE_\ell, \ \ell=6, 7                 & \ell &1      & 27, 56         & \text{minimal}       & \rD_{\ell-1}              & 2(\ell-1)      & \ell-1 \\
\rF_4,                                & 4    & 1     & 26             & \text{minimal}       & \rC_{3}                   & 6              & 3 \end{array}\]
 \caption{List of representations used in the proof of \cref{thm:DennisVaserstein}.} \label{table:dv-reps}
\end{table}
It is essential for the proof of~\cref{thm:DennisVaserstein} that the normalizer $N(A_{rs})$ is suffiently large.
The following lemma describes the ``obvious'' root subgroups which are contained $N(A_{rs})$ for arbitrary $R$.
\begin{lemma}\label{lemma:dv-normal} 
For every $\alpha \in \Delta_{r, s} \cup \Phi^+ $ one has $X_\alpha(R) \subseteq N(A_{rs})$. \end{lemma}
\begin{proof}
Notice that for every $i$ the group $\EP_i(R, I)$ is normalized by $\EP_i(R)$, hence it is normalized by $X_\alpha(R)$, $\alpha \in S_i^+$.
Since $\E(\Delta_{r, s}, R)$ normalizes $\U(\Sigma_r^- \cap \Sigma_s^-, I)$, we obtain the required assertion for $\alpha \in \Delta_{r, s}$.

@@

Notice that by the definition of $\Gamma$ and Matsumoto lemma~\cite[Lemma~2.3]{Ma69} for any $s\in I$, $ g\in B$ one has $x_{-\alpha_r}(s) \cdot g \cdot v^+ = g \cdot v^+$, therefore
\[ X_{-\alpha_{r}}(I)^{B} \subseteq \U(\Phi^-, I) \cap \Stab(v^+) \subseteq \U(\Delta_s^-, I) \subseteq \EP_s(R, I). \]
From the above inclusion we immediately obtain that
\begin{equation*} X_{\alpha_r}(I) \cdot X_{-\alpha_r}(I) \cdot B \cdot \EP_s(R, I) \subseteq X_{\alpha_r}(I) \cdot B \cdot \EP_s(R, I) \subseteq B \cdot \U(\Sigma_r^+, I) \cdot \EP_s(R, I) = B \cdot \EP_s(R, I), \end{equation*}
which together with the third statement of \cref{lemma:dv_unipotent} implies that
\begin{equation*} \label{rel:sred}
A = \U(\Phi^+\setminus\{\alpha_r\}, I) \cdot \U(\Phi^-, I) \cdot B \cdot \EP_s(R, I).
\end{equation*}
Finally, since $[B, X_{-\alpha_r}(R)] \subseteq \U(\Sigma_r^-, R) \cap \E(\Phi, R, I) = \U(\Sigma_r^-, I)$, we obtain the required assertion, indeed:
\begin{equation*} \label{rel:ninv} A^{X_{-\alpha_{r}}(R)} = \U(\Phi^+, I) \cdot \U(\Phi^-, I) \cdot B ^{X_{-\alpha_{r}}(R)} \cdot \EP_s(R, I) = A. \qedhere \end{equation*}
\end{proof}

\begin{lemma}\label{lemma:DVST}
Assume that $\Phi=\rC_\ell$, $s=1$, $r=\ell$ and that $\sr(I) \leq \ell-1$. Then $X_{-\alpha_1}(R) \subseteq N(A_{\ell 1})$.
\end{lemma}
\begin{proof}
Denote by $C$ the set consisting of elements $g \in \EP_1(R, I)$ for which matrix entries $(g_{i, 2})$, $i=2, \ldots, \ell$ form an $I$-unimodular column of height $\ell-1$.
Set $A' = \EP_\ell(R, I) \cdot \U(\Sigma_1^- \cap \Sigma_\ell^-, I) \cdot C$.
Applying \cref{item:asrUnipC} we can find for every $g \in \EP_1(R, I)$ some element $x \in \U(\Sigma_\ell^+ \cap \Delta_1, I)$ such that $xg \in C$.  
The equality $A_{\ell 1} = A'$ follows from this, indeed, for $g\in \EP_1(R, I)$ one has
\begin{equation*} \EP_\ell(R, I) \cdot \U(\Sigma_1^- \cap \Sigma_\ell^-, I) \cdot g \subseteq 
 \EP_\ell(R, I)x^{-1}  \cdot \U(\Sigma_1^-, I) \cdot xg \subseteq A'. \end{equation*}

@@

\section{Applications} \label{sec:applications}
\subsection{Subsystem factorizations}\label{sec:subsysfact}
Recall from~\cite{LNS11, V13} that a Chevalley group over a field $\G(\Phi, F)$ can be presented as a product of at most $3|\Phi^+|$ of its subgroups $\SL(2, F)$.
As an easy corollary of the relative Bass--Kolster decomposition we get that a similar factorization exists for relative groups over a rather general class of commutative rings.
\begin{thm}\label{thm:SL2width}
Let $I\trianglelefteq R$ be an ideal, $\Phi$ be an irreducible classical root system of rank $\ell$ satisfying one the following 4 conditions:
\[\begin{array}{l@{\quad}l@{\quad}l}
\Phi=\rA_\ell, & \sr(I)\leqslant 2, & N=3\left|\Phi^+\right| - \ell - 1; \\
\Phi=\rC_\ell, & \sr(I)\leqslant 3, & N=3\left|\Phi^+\right| - 2; \\
\Phi=\rB_\ell, \rD_\ell, & \asr(I)\leqslant 2, & N=4\left|\Phi^+\right| - 3\ell.
\end{array}\]
Then the principal congruence subgroup $\G(\Phi, R, I)$ can be presented as a product of at most $N$ of its subgroups, 
 where each subgroup is an isomorphic copy of $\SL(2, R, I)$.
\end{thm}

>>>>

\section{Introduction}\label{sec:intro}
This paper studies decompositions of Chevalley groups over commutative rings (see e.\,g.~\cite{VP} for the introduction to Chevalley groups over rings).
Recall that classical decompositions of Chevalley groups over fields such as Bruhat decomposition do not generalize to groups over more general rings.
Nevertheless, Chevalley groups over rings of {\it finite stable rank} do admit several remarkable ``parabolic decompositions'', i.\,e. decompositions formulated in terms of parabolic and unipotent subgroups.
The interest in these decompositions comes, in particular, from the study of stability problems for $\K_1$-functors modeled on Chevalley groups (see e.\,g.~\cite{St78, Si13}).

The first goal of this paper is to obtain analogues of these parabolic decompositions for congruence subgroups of Chevalley groups.
One of our main results is the following theorem, which is, essentially, a version of Dennis--Vaserstein decomposition for the relative elementary group $\E(\Phi, R, I)$
 (cf.~\cite[Theorems~2.5 and~4.1]{St78},~\cite[Theorem~1.2]{Si13}).
\begin{thm}\label{thm:DennisVaserstein}
Let $\Phi \neq \rE_8$ be an irreducible root system of rank $\ell\geqslant 2$ and $\{ r, s \}$ be a pair of distinct endnodes of the Dynkin diagram of $\Phi$.
Denote by $d$ the distance between $r$ and $s$ on the Dynkin diagram.
Assume that $\Phi$, $I$ and $\{r, s\}$ satisfy either of the following assumptions:
\begin{thmlist}
 \item $\sr(I) \leqslant d$ for $\Phi$ classical or $\Phi=\rG_2, \rF_4$;
 \item $\sr(I) \leqslant d$ for $\Phi=\rE_6,\rE_7$ with $\{r, s\} = \{2, \ell\}$;
 \item $\asr(I) \leqslant d$ for $\Phi=\rE_6,\rE_7$ with $\{r, s\} = \{1, \ell \}$, 
\end{thmlist}
Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$
\end{thm}
The notation for the groups and ring-theoretic invariants appearing in the statement of the above theorem is introduced in sections~\ref{sec:preliminaries} and~\ref{sec:stability-conditions}.
Other relative decompositions studied in this article are Gauss and Bass--Kolster decompositions (see~\cref{thm:srRI1,thm:BassKolster}).

As an application of parabolic factorizations, we deduce several {\it subgroup factorizations} of finite width for relative Chevalley groups.
For example, as a corollary of Bass--Kolster decomposition we obtain upper estimates of the number of factors in the presentation of a classical Chevalley group
 as a product of its subsystem subgroups $\SL(2, R, I)$, see~\cref{thm:SL2width}.
As an application of~\cref{thm:DennisVaserstein} we prove the following theorem inspired by the main result of~\cite{Nik07}.
\begin{thm}\label{thm:spin-sln-prod} Assume that $\sr(I) \leqslant 2$. Then the group $\Epin(2\ell, R, I)=\E(\rD_\ell, R, I)$ is a product of at most $9$ regularly embedded subgroups of type $\rA_{\ell-1}$. \end{thm}
As another application, we obtain results on the bounded generation of the relative elementary group.
Recall from~\cite[Theorem~2]{Va86} that $\E(\Phi, R, I)$ can be generated by the set of Tits--Vaserstein generators
 $z_\alpha(s, \xi) =  x_{-\alpha}(-\xi) x_\alpha(s) x_{-\alpha}(\xi)$, where $s\in I$ and $\xi\in R$
(in fact, a smaller generating set suffices, see~\cref{prop:Stepanov-theorem}).
As a further application of parabolic factorizations, we obtain in~\cref{thm:width} estimates of the width $\E(\Phi, R, I)$  
 with respect to this generating set in several special cases.

@@

Let $\G(\Phi, R)$ be the simply-connected Chevalley group of type $\Phi$ over an arbitrary commutative ring $R$ and let $\E(\Phi, R)$ be its elementary subgroup,
i.\,e. the subgroup generated by the elementary root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in R$, see~\cite{VP, St78, S}.

For an ideal $I \trianglelefteq R$ we denote by $\G(\Phi, R, I)$ the \emph{principal congruence subgroup} of level $I$.
Recall that the \emph{relative elementary Chevalley subgroup} $\E(\Phi, R, I) \leqslant \G(\Phi, R, I)$ is defined as the normal closure in $\E(\Phi, R)$ of the subgroup $\E(\Phi, I)$ generated by the set $\mathcal{X} = \{x_\alpha(\xi) \mid \alpha\in \Phi, \ \xi \in I \}$.

For a subset of roots $S\subseteq\Phi$ denote by $\mathcal{Z}(S)$ the union of $\mathcal{X}$ and $\{ z_\alpha(s, \xi) \mid s\in I, \ \xi \in R, \ \alpha \in S\}.$
The following result describes generating sets of $\E(\Phi, R, I)$.
\begin{prop}[{\cite[Theorem~3.4]{S}}] \label{prop:Stepanov-theorem} Let $\Phi$ be an irreducible root system of rank $\geq 2$ and $S \subseteq \Phi$ be an arbitrary parabolic subset of roots with special part $\Sigma_S$.
 Then the relative elementary subgroup $\E(\Phi, R, I)$ is generated as an abstract group by $\mathcal{Z}(\Sigma_S)$. \end{prop}
Denote by $I^{\indexbox{2}}$ the ideal of $R$ generated by all squares $a^2$, where $a\in I$.
 \begin{lemma}[{\cite[Corollary~3.3]{S}}]\label{lemma:Stepanov-ideal}
Let $\Phi$ be a root system of rank $\geqslant2$, let $R$ be a commutative ring and $I\trianglelefteq R$ be its ideal.
If $\Phi\neq\rC_\ell$ then $\E\left(\Phi, R, I^2\right)\leqslant\E(\Phi, I)$, otherwise $\E\left(\Phi, R, II^{\indexbox{2}}\right)\leqslant\E(\Phi, I)$.
\end{lemma}
For $\varepsilon\in R^*$ set $w_\alpha(\varepsilon) = x_\alpha(\varepsilon) x_{-\alpha}(-\varepsilon^{-1}) x_{\alpha}(\varepsilon).$
If $\rk(\Phi)\geqslant 2$ the following relation holds:
\begin{equation}\label{rel:R3}
w_\alpha(\varepsilon) x_{\beta}(\xi) w_\alpha(\varepsilon)^{-1} =
x_{\sigma_{\alpha}\beta} \left(\eta_{\alpha, \beta}\cdot\varepsilon^{-\langle\beta, \alpha \rangle}\xi\right), \quad \varepsilon\in R^*, \ \xi\in R.
\end{equation}
where $\eta_{\alpha, \beta}=\pm 1$. The coefficients $\eta_{\alpha, \beta}$ can be expressed in terms of the structure constants of the corresponding Lie algebra (see~\cite[\S13]{VP}).
For a reductive subset $\Delta \subseteq \Phi$ denote by $\widetilde{W}(\Delta)$ the \emph{extended Weyl group}, i.\,e. the subgroup of $\E(\Phi, R)$ generated by all $w_{\alpha}(1)$, $\alpha \in \Delta$.
\begin{lemma} \label{lemma:weylfacts} Let $\Phi$ be an irreducible root system and let $I$ be an ideal of $R$. 
\begin{lemlist}
\item \label{item-trans1} For every two roots $\alpha, \beta \in \Phi$ of the same length there exists $w \in \widetilde{W}(\Phi)$ such that $X_{\alpha}(I)^w = X_\beta(I)$.
\item \label{item-trans2} Let $\alpha_s\in \Pi$ be a fundamental root and $\alpha, \beta \in \Sigma^\pm_s$ be two roots having the same length and $s$-shape.
 Then there exists $w\in \widetilde{W}(\Delta_s)$ such that $X_\alpha(I)^w = X_\beta(I)$. \end{lemlist}
\end{lemma}
\begin{proof}
The first assertion follows from \eqref{rel:R3} and the well-known fact that $W(\Phi)$ acts transitively on the set of roots of the same length.
The second assertion follows from \eqref{rel:R3} and \cref{lemma:abs}.
\end{proof}

@@

From Levi decomposition~\eqref{rel:Levi-decomp} we obtain the following equality of subsets of $\E(\Phi, R, I)$:
\begin{multline*}
\U(\Phi^+, I)\cdot \U(\Phi^-, I) \cdot \E(\Delta_r, R, I) \cdot \EP_s(R, I) = \\
= \U(\Sigma_r^+, I)\cdot \U(\Sigma^-_r, I) \cdot \E(\Delta_r, R, I) \cdot \EP_s(R, I) = \hspace{5em} \\
\hspace{5em} = \EP_r(R, I) \cdot \E(\Delta_s, R, I) \cdot \U(\Sigma_s^-, I)\cdot \U(\Sigma_s^+, I) = \\
= \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).
\end{multline*}
Denote the above subset by $A_{rs}$. 
Notice that the equality $A_{rs} = \E(\Phi, R, I)$ implies symmetric equality $A_{sr} = \E(\Phi, R, I)$, 
 therefore it suffices to consider only the following possibilities for $\Phi$, $s$, $r$.

\begin{table}[htb]
\[\begin{array}{l @{\qquad} l @{\qquad} c @{\quad} c @{\quad} c @{\qquad} c @{\qquad} c @{\qquad} c}
\Phi                                  & s    &r      & |\Lambda(\pi)| & \text{type of $\pi$} & \text{type of $\Delta_r$} & |\Lambda(\pi')|& |\Gamma|  \\ \hline\vphantom{\Bigl(}
\rA_\ell, \ \ell\geqslant 2           & 1    &\ell   & \ell+1         & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rB_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell+1        & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rC_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\
\rD_\ell, \ \ell\geqslant 4           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 2  \\ 
\rD_\ell, \ \ell\geqslant 4           & \ell &\ell-1 & 2^{\ell-1}     & \text{half-spinor}   & \rA_{\ell-1}              & \ell           & \ell-2  \\
\rE_\ell, \ \ell=6, 7                 & \ell &2      & 27, 56         & \text{minimal}       & \rA_{\ell-1}              & \ell           & 3       \\ 
\rE_\ell, \ \ell=6, 7                 & \ell &1      & 27, 56         & \text{minimal}       & \rD_{\ell-1}              & 2(\ell-1)      & \ell-1 \\
\rF_4,                                & 4    & 1     & 26             & \text{minimal}       & \rC_{3}                   & 6              & 3 \end{array}\]
 \caption{List of representations used in the proof of \cref{thm:DennisVaserstein}.} \label{table:dv-reps}
\end{table}
For the proof~\cref{thm:DennisVaserstein} we need to show that the normalizer $N(A_{rs})$ is a sufficiently large subgroup.
This is accomplished in the series of lemmas below. For example, the following lemma describes root subgroups 
 contained in $N(A_{rs})$ regardless of whether the stability conditions of~\cref{thm:DennisVaserstein} are satisfied.
\begin{lemma}\label{lemma:dv-normal} 
For every $\alpha \in \Delta_{r, s} \cup \Phi^+ $ one has $X_\alpha(R) \subseteq N(A_{rs})$. \end{lemma}
\begin{proof}
Notice that for every $i$ the group $\EP_i(R, I)$ is normalized by $\EP_i(R)$, hence it is normalized by $X_\alpha(R)$, $\alpha \in S_i^+$.
Since $\E(\Delta_{r, s}, R)$ normalizes $\U(\Sigma_r^- \cap \Sigma_s^-, I)$, we obtain the required assertion for $\alpha \in \Delta_{r, s}$.

@@

Notice that by the definition of $\Gamma$ and Matsumoto lemma~\cite[Lemma~2.3]{Ma69} for any $s\in I$, $ g\in B$ one has $x_{-\alpha_r}(s) \cdot g \cdot v^+ = g \cdot v^+$, therefore
\[ X_{-\alpha_{r}}(I)^{B} \subseteq \U(\Phi^-, I) \cap \Stab(v^+) \subseteq \U(\Delta_s^-, I) \subseteq \EP_s(R, I). \]
From the above inclusion we immediately obtain that
\begin{equation*} X_{\alpha_r}(I) \cdot X_{-\alpha_r}(I) \cdot B \cdot \EP_s(R, I) \subseteq X_{\alpha_r}(I) \cdot B \cdot \EP_s(R, I) \subseteq B \cdot \U(\Sigma_r^+, I) \cdot \EP_s(R, I) = B \cdot \EP_s(R, I), \end{equation*}
which together with the third statement of \cref{lemma:dv_unipotent} implies that
\begin{equation*} \label{rel:sred}
A = \U(\Phi^+\setminus\{\alpha_r\}, I) \cdot \U(\Phi^-, I) \cdot B \cdot \EP_s(R, I).
\end{equation*}
Finally, since $[B, X_{-\alpha_r}(R)] \subseteq \U(\Sigma_r^-, R) \cap \E(\Phi, R, I) = \U(\Sigma_r^-, I)$, we obtain the required assertion, indeed:
\begin{equation*} \label{rel:ninv} A^{X_{-\alpha_{r}}(R)} = \U(\Phi^+, I) \cdot \U(\Phi^-, I) \cdot B ^{X_{-\alpha_{r}}(R)} \cdot \EP_s(R, I) = A. \qedhere \end{equation*}
\end{proof}
We also need a separate lemma to deal with the symplectic case.
\begin{lemma}\label{lemma:DVST}
Assume that $\Phi=\rC_\ell$, $s=1$, $r=\ell$ and that $\sr(I) \leq \ell-1$. Then $X_{-\alpha_1}(R) \subseteq N(A_{\ell 1})$.
\end{lemma}
\begin{proof}
Denote by $C$ the set consisting of elements $g \in \EP_1(R, I)$ for which matrix entries $(g_{i, 2})$, $i=2, \ldots, \ell$ form an $I$-unimodular column of height $\ell-1$.
Set $A' = \EP_\ell(R, I) \cdot \U(\Sigma_1^- \cap \Sigma_\ell^-, I) \cdot C$.
Applying \cref{item:asrUnipC} we can find for every $g \in \EP_1(R, I)$ some element $x \in \U(\Sigma_\ell^+ \cap \Delta_1, I)$ such that $xg \in C$.  
The equality $A_{\ell 1} = A'$ follows from this, indeed, for $g\in \EP_1(R, I)$ one has
\begin{equation*} \EP_\ell(R, I) \cdot \U(\Sigma_1^- \cap \Sigma_\ell^-, I) \cdot g \subseteq 
 \EP_\ell(R, I)x^{-1}  \cdot \U(\Sigma_1^-, I) \cdot xg \subseteq A'. \end{equation*}

@@

\section{Applications} \label{sec:applications}
\subsection{Subsystem factorizations}\label{sec:subsysfact}
Recall from~\cite{LNS11, V13} that a Chevalley group over a field $\G(\Phi, F)$ can be presented as a product of at most $3|\Phi^+|$ of its subgroups $\SL(2, F)$.
As an easy corollary of Bass--Kolster decomposition we get that a similar factorization also holds for relative groups over rings of small stable rank.
\begin{thm}\label{thm:SL2width}
Let $I\trianglelefteq R$ be an ideal, $\Phi$ be an irreducible classical root system of rank $\ell$ satisfying one the following conditions:
\[\begin{array}{l@{\quad}l@{\quad}l}
\Phi=\rA_\ell, & \sr(I)\leqslant 2, & N=3\left|\Phi^+\right| - \ell - 1; \\
\Phi=\rC_\ell, & \sr(I)\leqslant 3, & N=3\left|\Phi^+\right| - 2; \\
\Phi=\rB_\ell, \rD_\ell, & \asr(I)\leqslant 2, & N=4\left|\Phi^+\right| - 3\ell.
\end{array}\]
Then the principal congruence subgroup $\G(\Phi, R, I)$ can be presented as a product of at most $N$ of its subgroups, 
 where each subgroup is an isomorphic copy of $\SL(2, R, I)$.
\end{thm}
